Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-18 paper/paper.tex
Stuff

>>>>

\section{Conclusions and Future Work}

Hylomorphisms are a general recursion scheme that can encode any other
recursion scheme, and that satisfy a number of algebraic laws that can be used
to reason about program equivalences. Not many
pieces of work formalise them in a proof assistant, partly due to the
difficulty of dealing with termination, and reasoning about functional
extensionality. In this work, we tackle these problems in a \emph{fully
axiom-free} way that targets the extraction of idiomatic OCaml code.  This
formalisation allows the use of program calculation techniques in Coq to derive
formally optimised implementations from naive specifications. Furthermore, the
rewritings that are applied to specifications are formal, machine-checked
proofs that the resulting program is extensionally equal to the input
specification.

Remaining axiom-free forces us to deal with the well-known
\emph{setoid hell}. As part of the future improvements, we will study how to
mitigate this problem. At the moment, we use a short ad-hoc tactic that is able
to automatically discharge many of these proofs in simple settings.  We will
study the more thorough and systematic use of proof automation for respectful
morphisms.  Generalisd rewriting in proofs involving setoids tends to be quite
slow, due to the large size of the terms that need to be rewritten. Sometimes,
this size is hidden in implicit arguments and coercions. We will study
alternative formulations to try to improve the performance of the rewriting
tactics (e.g.  canonical structures).  Currently, Coq is unable to inline a
number of trivially inlineable definitions.  We will study alternative
definitions, or extensions to Coq's code extraction mechanisms to force the
full inlining of all container code that is used in hylomorphisms.  Finally,
proving termination still remains a hurdle. In our framework this reduces to
proving that the anamorphism
terminates in all inputs, and we provide a convenient connection to
well-founded recursion. Furthermore, recursive coalgebras compose with natural
transformations, which allows the reuse
of a number of core recursive coalgebras. A possible interesting future line of work
is the use of the approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to
improve code extraction from divide-and-conquer computations whose termination
does not require an external proof.
